Binning Strategies and Related Loss for Large Data

### 2015-05-12 Paper.Rnw
Fix appendix notation to match paper

While bin widths can be chosen as any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual artefacts in the binned scatterplot. For example, if the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins in the binned scatterplot.  These gaps of white space are undesirable because they create visual discontinuity that interferes with the interpretation of bivariate density. 
\hh{The grid of white spaces also creates an optical illusion, the Hermann-grid illusion~\citep{hermann:1870, spillmann:1994}, which makes dark spots appear in the crossings of the lines and additionally interferes with an evaluation.}

@@

Figure~\ref{fig:StandardFreqBinning4710none} displays several spatially binned scatterplots. On the right frequency is mapped continuously to colour; on the left  there is a varying numbers of  standard binned frequencies. It is trivial to discern differences between similarly shaded tiles
when there are only four shades. This becomes more difficult -- but still possible -- at seven frequency bins, but at ten frequency bins we are hardly able to discriminate between colours better than in the continuous shading. 
This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. We will therefore use an upper limit of ten frequency bins in our exploration of frequency loss.

@@

The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} bin counts from the same sets of simulated data. 
Frequency binning using log counts exhibits a very similar behaviour to binning of untransformed frequencies.
The difference in scales makes it impossible to directly compare frequency loss and log frequency loss. 
But the same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.

@@

The following is proof that when univariate data is uniformly distributed at resolution $\alpha$ and standard rectangular binning is used with binwidths that are a scalar multiple of the data resolution (i.e. $\omega = k\alpha$ for some $k \in \{1,2,\dots\})$, then spatial loss is minimized by setting the binning origin to $\alpha/2$ units below the minimum data value; set $\beta = x_{(1)} - \alpha/2$.\\

\noindent Let $x_1, x_2, \dots, x_k \in \mathbb{R}$ represent the values in a single bin such that $x_{i+1} = x_i + \alpha$ for some constant $\alpha \in \mathbb{R}$. Thus $x_j = x_1 + (j-1)\alpha$.\\

\noindent Suppose then that we bin the data using standard rectangular binning with origin, $\beta = x_1 - \theta$, and binwidth $\omega$; where $\theta$ is the \textit{origin offset} from the data. Thus $b(x_j) = \beta + \omega/2 = (x_1 - \theta) + (k\alpha/2)$\\

@@

$ b(x_i) = Q_x(.5) $ \\
$ \Rightarrow (x_1 - \theta) + (k\alpha/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
$ \Rightarrow 2x_1 - 2\theta + k\alpha = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
$ \Rightarrow -2\theta + k\alpha = (\lceil\frac{k+1}{2}\rceil - 1)\alpha  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha  $ \\
$ \Rightarrow -2\theta + k\alpha = ((k+1) - 2)\alpha  $ \\
$ \Rightarrow -2\theta = -\alpha $ \\
$ \Rightarrow \theta = \alpha/2 $ \\

Thus the optimal offset for reducing spatial loss in this scenario is $\theta = \alpha/2$.  This result holds for data that is symmetrically distributed within the bin since the median will not change.  It extends to multiple contiguous bins with resolution $\alpha$ data that has symmetrically distributed data withing each bin. \\

>>>>

While bin widths can be chosen as any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual artifacts in the binned scatterplot. For example, if the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins in the binned scatterplot.  These gaps of white space are undesirable because they create visual discontinuity that interferes with the interpretation of bivariate density. 
\hh{The grid of white spaces also creates an optical illusion, the Hermann-grid illusion~\citep{hermann:1870, spillmann:1994}, which makes dark spots appear in the crossings of the lines and additionally interferes with an evaluation.}

@@

Figure~\ref{fig:StandardFreqBinning4710none} displays several spatially binned scatterplots. On the right frequency is mapped continuously to color; on the left  there is a varying numbers of  standard binned frequencies. It is trivial to discern differences between similarly shaded tiles
when there are only four shades. This becomes more difficult -- but still possible -- at seven frequency bins, but at ten frequency bins we are hardly able to discriminate between colors better than in the continuous shading. 
This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. We will therefore use an upper limit of ten frequency bins in our exploration of frequency loss.

@@

The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} bin counts from the same sets of simulated data. 
Frequency binning using log counts exhibits a very similar behavior to binning of untransformed frequencies.
The difference in scales makes it impossible to directly compare frequency loss and log frequency loss. 
But the same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.

@@

\ktm{The following proof assumes three conditions for standard rectangular binning of a set univariate data hold: (i) data are recorded to a resolution of $\alpha_x$ units in the $X$ dimension, (ii) points are symmetric distributed within standard rectangular bins, (iii) the bin width, $\omega_x$, is an integer multiples of $\alpha_x$ (i.e. $\omega_x = k\alpha_x$ for some $k \in \{1,2,\dots\}$).  Under these conditions it will be shown that spatial loss is minimized by setting the binning origin, $\beta_x$, to $\alpha_x/2$ units below the minimum data value (i.e. $\beta_x = x_{(1)} - \alpha_x/2$).\\ }

\noindent Let $x_1, x_2, \dots, x_k \in \mathbb{R}$ represent the values in a single bin such that $x_{i+1} = x_i + \alpha_x$ for some constant $\alpha_x \in \mathbb{R}$. Thus $x_j = x_1 + (j-1)\alpha_x$.\\

\noindent Suppose then that we bin the data using standard rectangular binning with origin, $\beta_x = x_1 - \theta$, and binwidth $\omega$; where $\theta$ is the \textit{origin offset} from the data. Thus $b(x_j) = \beta_x + \omega/2 = (x_1 - \theta) + (k\alpha_x/2)$\\

@@

$ b(x_i) = Q_x(.5) $ \\
$ \Rightarrow (x_1 - \theta) + (k\alpha_x/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
$ \Rightarrow 2x_1 - 2\theta + k\alpha_x = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha_x ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha_x ) $ \\
$ \Rightarrow -2\theta + k\alpha_x = (\lceil\frac{k+1}{2}\rceil - 1)\alpha_x  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha_x  $ \\
$ \Rightarrow -2\theta + k\alpha_x = ((k+1) - 2)\alpha_x  $ \\
$ \Rightarrow -2\theta = -\alpha_x $ \\
$ \Rightarrow \theta = \alpha_x/2 $ \\

Thus the optimal offset for reducing spatial loss in this scenario is $\theta = \alpha_x/2$.  This result holds for data that is symmetrically distributed within the bin since the median will not change.  It extends to multiple contiguous bins with resolution $\alpha_x$ data that has symmetrically distributed data withing each bin. \\
